---
title: "Linear regression"
author: "Łukasz Wawrowski"
output: html_document
---

The purpose of regression analysis is to determine, on the basis of the available information, the unknown value of the analyzed variable. We know the sales values and the number of customers in a given Rossmann store and would like to determine the possible level of sales for a given number of customers, e.g. 1000 customers. 

Similarly as in the correlation analysis, the starting point in a simple regression is to create a scatter plot. In order to apply the regression model, there must be a correlation relationship between the variables and it must be a linear relationship. Then we have to determine the dependent variable ($y$) and the independent variable ($x$).

Consider a simple example of income and spendings:

```{r include=FALSE}
knitr::opts_chunk$set(cache = FALSE, echo = FALSE, message = F, error = F)
```


```{r}
library(tidyverse)

options(scipen = 999)

load("data/dane.RData")

rossmann_1 <- rossmann %>% 
  filter(sklep_id == 1, czy_otwarty == "Tak")

d <- data.frame(spendings=c(2300,1800,2400,2300,2800,2000,2100),
                income=c(2600,2400,2900,2800,3000,2500,2700))

d %>% knitr::kable()
```

The relationship between spendings and income seems obvious - for $y$ we will take spendings and $x$ is income. In the chart, this dependence is as follows:

```{r}
ggplot(d, aes(x = income, y = spendings)) + 
  geom_point(size = 2) +
  xlab("income (X)") + 
  ylab("spendings (Y)") +
  xlim(2400,3000) +
  ylim(1600,3100) +
  theme_light()
```

We are interested in creating a model that simplifies reality to the level of a formula for a simple line, whose general equation is as follows:

$$y=a_1\cdot x+a_0$$

For only two points, determining the $a_1$ and $a_0$ coefficients would not be a problem. However, for the given example, the Ordinary Least Squares (OLS) method should be used, in which we minimize the distance of points from the line. 

Now let's try to fit a few lines.

```{r}
ggplot(d, aes(x = income, y = spendings)) + 
  geom_point(size = 2) +
  geom_hline(yintercept = 2243, color = "blue", alpha = 0.8, size = 1.1) +
  geom_abline(slope = 1.9, intercept = -2780, color = "green", alpha = 0.8, size = 1.1) +
  geom_abline(slope = 1.357, intercept = -1421.429, color = "red", alpha = 0.8, size = 1.1) +
  xlab("income (X)") + 
  ylab("spendings (Y)") +
  xlim(2400,3000) +
  ylim(1600,3100) +
  theme_light()
```

In the next step we calculate the differences between the existing points and the corresponding values on the line: 

```{r}
d <- d %>% 
  mutate(niebieska_y=income-2243,
         zielona_y=1.9*income-2780,
         czerwona_y=1.357*income-1421.429) %>% 
  mutate(blue=(spendings-niebieska_y)^2,
         green=(spendings-zielona_y)^2,
         red=(spendings-czerwona_y)^2)

line1 <- data.frame(x = c(d$income[1], d$income[1]), y=c(d$spendings[1], d$czerwona_y[1]))
line2 <- data.frame(x = c(d$income[2], d$income[2]), y=c(d$spendings[2], d$czerwona_y[2]))
line3 <- data.frame(x = c(d$income[3], d$income[3]), y=c(d$spendings[3], d$czerwona_y[3]))
line4 <- data.frame(x = c(d$income[4], d$income[4]), y=c(d$spendings[4], d$czerwona_y[4]))
line5 <- data.frame(x = c(d$income[5], d$income[5]), y=c(d$spendings[5], d$czerwona_y[5]))
line6 <- data.frame(x = c(d$income[6], d$income[6]), y=c(d$spendings[6], d$czerwona_y[6]))
line7 <- data.frame(x = c(d$income[7], d$income[7]), y=c(d$spendings[7], d$czerwona_y[7]))

ggplot(d, aes(x = income, y = spendings)) + 
  geom_point(size = 2) +
  geom_hline(yintercept = 2243, color = "blue", alpha = 0.8, size = 1.1) +
  geom_abline(slope = 1.9, intercept = -2780, color = "green", alpha = 0.8, size = 1.1) +
  geom_abline(slope = 1.357, intercept = -1421.429, color = "red", alpha = 0.8, size = 1.1) +
  geom_line(data = line1, aes(x=x, y=y), color = "grey60", size = 1.2) +
  geom_line(data = line2, aes(x=x, y=y), color = "grey60", size = 1.2) +
  geom_line(data = line3, aes(x=x, y=y), color = "grey60", size = 1.2) +
  geom_line(data = line4, aes(x=x, y=y), color = "grey60", size = 1.2) +
  geom_line(data = line5, aes(x=x, y=y), color = "grey60", size = 1.2) +
  geom_line(data = line6, aes(x=x, y=y), color = "grey60", size = 1.2) +
  geom_line(data = line7, aes(x=x, y=y), color = "grey60", size = 1.2) +
  xlab("income (X)") + 
  ylab("spendings (Y)") +
  xlim(2400,3000) +
  ylim(1600,3100) +
  theme_light()
```

By denoting $y_i$ as the actual value of spendings and $\hat{y_i}$ as the value lying on the line we want to minimize the expression $\sum\limits_{i=1}^{n}{(y_{i}-\hat{y}_{i})^2} \rightarrow min$. The difference $y_{i}-\hat{y}_{i}$ is called residual. By calulating these values for the analyzed lines, we will get the following results:

```{r}
d %>% 
  select(6:8) %>% 
  pivot_longer(1:3) %>% 
  group_by(name) %>% 
  summarise(sum_squares=round(sum(value))) %>% 
  ungroup() %>% 
  rename(line=name) %>% 
  arrange(sum_squares) %>% 
  knitr::kable()
```

As we can see, the smallest value of the sum of squares is observed for the red line. We are now interested in the equation of this line. Assuming the earlier symbols, the general form of the regression line is as follows:

$$\hat{y}_{i}=a_{1}x_{i}+a_{0}$$

where $y$ with a hat ($\hat{y}$ is a theoretical value, lying on a line. 

Therefore the empirical/real values (y) will be described by the formula:

$$y_{i}=a_{1}x_{i}+a_{0}+u_{i}$$

where $u_i$ is the residual calculated as $u_{i}=y_{i}-\hat{y}_{i}$. 

In the Rossmann case the dependent variable will be the sales ($y$), which will be explained by the number of customers ($x$). Our goal is to find a formula of a straight line that will be as close as possible to all points of the plot. We need to determine the slope ($a_1$) and the point of intersection with the OY axis ($a_0$) - intercept or constant.

```{r ross-reg, echo=FALSE, fig.height=5, fig.width=8, warning=FALSE, message=FALSE}

ggplot(rossmann_1, aes(y=sprzedaz, x=liczba_klientow)) + 
  geom_smooth(method = "lm", se = F, colour="grey80") + 
  geom_point() +
  ylab("Sales") + xlab("Number of clients") +
  theme_bw()

```

The values of these coefficients can be calculated using the following formula:

$$
a_{1}=\frac{\sum\limits_{i=1}^{n}{(x_{i}-\bar{x})(y_{i}-\bar{y})}}{\sum\limits_{i=1}^{n}{(x_{i}-\bar{x})^{2}}}
$$

or knowing Pearson's linear correlation coefficient:

$$a_{1}=r\frac{S_{y}}{S_{x}}$$

The value of the intercept can be obtained from the formula:

$$a_{0}=\bar{y}-a_{1}\bar{x}$$

where:

- $r$ - Pearson linear correlation coefficient for $x$ and $y$,
- $S_y$ - standard deviation of $y$,
- $S_x$ - standard deviation of  $x$,
- $\bar{y}$ - an average of $y$,
- $\bar{x}$ - an average of $x$.

On this basis, we determine that the straight line we are interested in has the following formula:

$$\hat{y}_i=10,45x_i-1091,22$$

A slope ($a_1$) indicates how much the average value of the dependent variable (y) will change when the value of the independent variable (x) increases by a unit. In other words: as the value of the x variable increses by 1 the value of the y increses by a slope. 

In our example, an increase in the number of customers by 1 person will result in an average sales increase of 10.45 EUR. 

The intercept ($a_0$) is the value of the dependent variable (y), if the value of the independent variable (x) is 0. Special care should be taken when interpreting this coefficient, because it is often meaningless. If x variable never is equal to 0, there is no interest in the intercept.

In the analyzed example, the $a_0$ coefficient indicates that with zero customers, sales in store 1 will be -1091,22 euros, which have no sense.
